Slitasje og kjørespor

### 2022-12-16 DRAFT_slitasjeOgKjørespor.Rmd
plan

```{r, eval=F}
infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
```

It's easuer to see what's happeing if we zoom in a bit. 
Lets get a boundary box around Trondheim.
```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(infra))
```

@@

```{r, fig.cap="Infrastrcture indeks over Trondheim. Major roads are in white."}
tm_shape(infra_trd)+
  tm_raster(title="Infrastructure Index",
            style="cont",
            palette = "viridis")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")
```

@@

#### Vegetation zones (bioregions)

I want to look at wheter the indicator values can be averages over
smaller and more meaningful areas (ecological homogeneous areas) than
the 5 default regions. One option is this data set with vegetation zone.

```{r, eval=F}
path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp",
               "/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp")
```

```{r, eval=F}
soner <- sf::read_sf(path)
```

```{r, eval=F}
tm_shape(soner)+
  tm_polygons(col = "NAVN",
              title = "Vegetation zone")
```

It could make sense to combine some zones, but I will later split the
zones by region, and then merge zones that are very uncommon in a
specific region.

#### Regions

Importing a shape file with the regional delineation.

```{r}
#reg <- sf::st_read("data/regions.shp")
#st_crs(reg)
```

@@

I can scale the indicator for each polygon, or I can chose to aggregate
them first. If the scaled value is representative and precise at the
polygon level, then I could scale at that level. However, the combined
surveyed area is still so little of the total area of Norway, so that
only producing indicator values for these leaves the indicator without
much value for regional assessments. Later when we produce assessments
and indicator values at the regional level, we cannot simply do an area
weighting of the polygons in each region. This is because we don't want
to assume that the polygons are representative far outside of the mapped
area. But perhaps we can assume them to be representative inside
*homogenous ecological areas*.

An option is to export one map with just the raw un-scaled slitasje
variable at the polygon level, and another, scaled indicator map
aggregated to either

1.  each *dekningskart* (map of surveyed area)

2.  each vegetation zone (e.g. boreonemal zone in Østlandet; excluding
    zones with too little data points)

3.  each vegetation region (aka bioregion, e.g. boreonemal zone and
    continental section in Østlandet)

4.  each region cross tabulated agains infrastructure index classes

5.  each pixel of a give size, say 1 x 1 km

Option 5 is problematic because we often will have too few points to be
able to estimate credible mean. The second and third option can be good
for certin biotic variables that depends on local species pools ect.
However, slitasje is perhaps best explained by human presence, and
option 4 therefore sounds reasonable.

@@

>>>>

Import a stars proxy (no data imported yet)
```{r, eval=F}
infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
```

It's easier to see what's happening if we zoom in a bit. 
Lets get a boundary box around Trondheim.
```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(infra))
```

@@

```{r, fig.cap="Infrastrcture indeks over Trondheim. Major roads are in white."}
map_trd <- tm_shape(infra_trd)+
  tm_raster(title="Infrastructure Index",
            style="cont",
            palette = "viridis")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")
map_trd
```

@@

```{r}
names(infra_trd) <- "infrastructureIndex"
infra_trd_reclassed <-  infra_trd %>%
  mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 9 ~ 2,
    infrastructureIndex >= 9 ~ 3
  ))
```

```{r}
map_trd_reclassed <- tm_shape(infra_trd_reclassed)+
  tm_raster(title="Infrastructure Index",
            #palette = "viridis",
            style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")

```{r, fig.cap="Infrastrcture indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
tmap_arrange(map_trd,
             map_trd_reclassed,
             ncol=1)
```

I tweaked the thresholds so that most of the forest next to Trondheim was i the second lowest class.

@@

I can scale the indicator for each polygon, or I can chose to aggregate
them first. If the scaled value is representative and precise at the
polygon level, then I could scale at that level. However, the combined
surveyed area is still so little of the total area of Norway, so that
only producing indicator values for these leaves the indicator without
much value for regional assessments. Later when we produce assessments
and indicator values at the regional level, we cannot simply do an area
weighting of the polygons in each region. This is because we don't want
to assume that the polygons are representative far outside of the mapped
area. But perhaps we can assume them to be representative inside
*homogenous ecological areas*. That's where the infrastructure index comes in. I will scale the indicators at the polygon level. I then take the simplified infrastructure index and, one class at the time, extract the indicator value that intersect with those cells, and extrapolate the mean of those values to all similar cell within a given region.

@@

<!-- Text here -->
Bootstrapping of nin-polygons per infrastructure class and region?
